Ship-based lidar evaluation of Southern Ocean low clouds in the storm-resolving general circulation model ICON and the ERA5 and MERRA-2 reanalyses

### 2025-07-18 manuscript_agu_rev1.tex
Various

\begin{keypoints}
\item Large cloud biases are present in the global storm-resolving model, but are lower in several key aspects than in the reanalyses.
\item The models tend to underestimate total cloud fraction, very low-level ($<$~250~m) cloud and fog, while overestimating cloud occurrence peak at 500 m.
\item A ``too few, too bright'' problem of underestimated cloud fraction, compensated by overestimated cloud albedo, is present in MERRA-2.
\end{keypoints}

@@

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds, temperature, and humidity profiles over the Southern Ocean in one such model, ICON, as well as two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, ICON relies entirely on explicit simulation of clouds instead of subgrid-scale parameterizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with ICON and reanalyses. We subsetted our results by cyclonic activity and stability. We found that ICON and reanalyses underestimate lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog or very low-level clouds are especially underestimated in the reanalyses. However, ICON and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level and lower stability. Southern Ocean cloud and fog biases are an important problem in the analyzed models and result in radiation balance biases.

@@

Some of the observatial data were likely used in reanalysis assimilation. The Macquarie Island station surface measurements and radiosonde profiles (not used in our analysis) were sent to the World Meteorological Organization Global Telecommunication System (GTS). The surface-level, lidar and radiosonde measurements on the RSV \textit{Aurora Australis} were not used outside of research purposes.

@@

For all observations considered (Fig.~\ref{fig:cloud-occurrence}a), the data show cloud occurrence peaking near the surface, whereas the models show a higher peak (at about 500~m). The models generally underestimate the total cloud fraction by 10--30\% and show a strong drop in cloud occurrence near the surface, which is not identified in the observations. ICON and ERA5 overestimate cloud occurrence at their peak (between 0 and 1~km). Above 1~km, ICON and MERRA-2 underestimate cloud occurrence, but ERA5 is accurate to about 3\% or less. The exaggerated peak in models is partly explained by the LCL distribution, which peaks about 300~m higher in the models than in the observations (near the surface), although this is not very pronounced. This is indicative of near-surface RH often being close to saturation in the observations but not in the models (Fig.~\ref{fig:stats-hist-surf}b). There are multiple possible reasons for this bias, such as how the statistical distribution of RH within a grid cell is represented in the models, the air--sea moisture flux parameterization, or weaker stability in the models can cause more boundary mixing across heights and thus lower near-surface RH.

@@

We analyzed a total of about 2400 days of lidar and 2300 radiosonde observations from 31 campaigns and the Macquarie Island subantarctic station, covering the Atlantic, Australian, and New Zealand sectors of the SO over 10 years. This dataset, together with the use of a ground-based lidar simulator, provided a comprehensive basis for evaluating SO cloud and thermodynamic profile biases in the GSRM ICON and the ERA5 and MERRA-2 reanalyses. Our analysis provides a unique evaluation perspective, complementary to satellite observations for evaluating boundary layer clouds and fog, which are predominant in this region. We did not, however, analyze the cloud phase based on ground-based observations. Cloud phase can have a strong impact on the SW radiative transfer due to larger and therefore less numerous hydrometeors (for the same amount of water) scatter much less SW radiation. Especially, the underestimation of fog or very low-level clouds is very substantial in the reanalyses, and we showed that this relates to cloud formation or persistence at RH between 80 and 100\% in the boundary layer in the observations, while in models RH values less than 100\% are associated with clear bins. We subsetted the dataset by low and high latitude SO bands, cyclonic activity, and stability in order to identify how these conditions influence the biases. The main identified biases are summarized in Table~\ref{tab:summary} and discussed below.

>>>>

\begin{keypoints}
\item Large cloud biases are present in the global storm-resolving model, but are lower in several key aspects than in the reanalyses.
\item The models tend to underestimate total cloud fraction and very low-level ($<$~250~m) cloud and fog, while overestimating cloud occurrence peak at 500 m.
\item A ``too few, too bright'' problem of underestimated cloud fraction, compensated by overestimated cloud albedo, is present in MERRA-2.
\end{keypoints}

@@

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds, temperature, and humidity profiles over the Southern Ocean in one such model, ICON, as well as two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, ICON relies entirely on explicit simulation of clouds instead of subgrid-scale parameterizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with ICON and reanalyses. We subset our results by cyclonic activity and stability. We found that ICON and reanalyses underestimate lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog or very low-level clouds are especially underestimated in the reanalyses. However, ICON and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level and lower stability. Southern Ocean cloud and fog biases are an important problem in the analyzed models and result in radiation balance biases.

@@

Some of the observatial data were likely used in reanalysis assimilation. The Macquarie Island station surface measurements and radiosonde profiles (not used in our analysis) were sent to the World Meteorological Organization Global Telecommunication System (GTS). The measurements on the RSV \textit{Aurora Australis} and HMNZS \textit{Wellington} were not used outside of research purposes. The AWS measurements but not lidar or radiosonde measurements on the RV \textit{Tangaroa} voyages were collected by the New Zealand MetService and communicated to the GTS. The ceilometer measurements on NBP1704 were not used outside of research purposes.

@@

For all observations considered (Fig.~\ref{fig:cloud-occurrence}a), the data show cloud occurrence peaking near the surface, whereas the models show a higher peak (at about 500~m). The models generally underestimate the total cloud fraction by 10--30\% and show a strong drop in cloud occurrence near the surface, which is not identified in the observations. ICON and ERA5 overestimate cloud occurrence at their peak (between 0 and 1~km). Above 1~km, ICON and MERRA-2 underestimate cloud occurrence, but ERA5 is accurate to about 3\% or less. The exaggerated peak in models is partly explained by the LCL distribution, which peaks about 300~m higher in the models than in the observations (near the surface), although this is not very pronounced. This is indicative of near-surface RH often being close to saturation in the observations but not in the models (Fig.~\ref{fig:stats-hist-surf}b). There are multiple possible reasons for this bias, such as how the statistical distribution of RH within a grid cell is represented in the models, the air--sea moisture flux parameterization, or weaker stability in the models, which can cause more boundary mixing across heights and thus lower near-surface RH.

@@

We analyzed a total of about 2400 days of lidar and 2300 radiosonde observations from 31 campaigns and the Macquarie Island subantarctic station, covering the Atlantic, Australian, and New Zealand sectors of the SO over 10 years. This dataset, together with the use of a ground-based lidar simulator, provided a comprehensive basis for evaluating SO cloud and thermodynamic profile biases in the GSRM ICON and the ERA5 and MERRA-2 reanalyses. Our analysis provides a unique evaluation perspective, complementary to satellite observations for evaluating boundary layer clouds and fog, which are predominant in this region. We did not, however, analyze the cloud phase based on ground-based observations. Cloud phase can have a strong impact on the SW radiative transfer due to larger and therefore less numerous hydrometeors (for the same amount of water) scatter much less SW radiation. Especially, the underestimation of fog or very low-level clouds is very substantial in the reanalyses, and we showed that this relates to cloud formation or persistence at RH between 80 and 100\% in the boundary layer in the observations, while in models RH values less than 100\% are associated with clear bins. We subset the dataset by low and high latitude SO bands, cyclonic activity, and stability in order to identify how these conditions influence the biases. The main identified biases are summarized in Table~\ref{tab:summary} and discussed below.
